**Best practices for applying machine learning to bacterial 16S rRNA gene sequencing data**

### 2019-07-15 submission/manuscript.Rmd
Sent to Pat

---
title: '**Evaluation of machine learning methods for 16S rRNA gene data**'
csl: mbio.csl
fontsize: 11pt
output:
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
  word_document: default
geometry: margin=1.0in
bibliography: references.bib
---

@@

Begüm D. Topçuoğlu${^1}$, Nick Lesniak${^1}$, Jenna Wiens${^2}$, Mack Ruffin${^3}$, Patrick D. Schloss^1$\dagger$^

@@

Machine learning (ML) modeling of the human microbiome has the potential to identify the microbial biomarkers and aid in diagnosis of many chronic diseases such as inflammatory bowel disease, diabetes and colorectal cancer. Progress has been made towards developing ML models that predict health outcomes from bacterial abundances, but rigourous ML models are scarce due to the flawed methods that call the validity of developed ML models into question. Furthermore, the use of black box ML models has hindered the validation of microbial biomarkers. To overcome these challenges, we benchmarked seven different ML models that use fecal 16S rRNA sequences to predict the presence/absence of colorectal cancer (CRC) lesions (n=490 patients, 261 controls and 229 cases). To show the effect of model selection, we assessed the predictive performance, interpretability, and computational efficiency of the following models: L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest, and extreme gradient boosting (XGBoost). The random forest model was best at detecting CRC lesions with an AUROC of 0.695 but it was slow to train (83.2 h) and hard to interpret. Despite its simplicity, L2-regularized logistic regression followed random forest in predictive performance with an AUROC of 0.680, and it trained much faster (12 min). In this study, we showed that ML models should be chosen based on expectations of predictive performance, interpretability and available computational resources. Additionally, we established standards for the development of modeling pipelines for microbiome-associated ML models

@@

Prediction of health outcomes using ML is rapidly being adopted by human-associated microbiome studies. However, the developed ML models so far are overoptimistic in terms of validity and predictive performance. Without rigorous ML pipelines, we cannot trust ML models. Before we can speed up progress, we need to slow down, define and start using good ML practices.
\newpage

@@

Advances in sequencing technology and decreasing costs of generating 16S rRNA gene sequences have allowed rapid exploration of human associated microbiome and its health implications. Currently, the human microbiome field is growing at an unprecedented rate and as a result, there is an increasing demand for methods that identify associations between members of the microbiome and human health. However, this is difficult as human associated microbial communities are remarkably complex, high-dimensional and uneven within and between individuals with the same disease. It is unlikely that a single species can explain a disease. Instead, subsets of those communities, in relation to one another and to their host, account for differences in health outcomes.

Machine learning (ML) methods are effective at recognizing and highlighting patterns in complex microbial datasets. They learn from existing data to predict the outcomes of new data and allow us to infer on the reasons underlying that prediction.  Therefore, researchers have started to explore the utility of ML models that use microbiota associated biomarkers to predict human health and to understand the ecological basis of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases (IBD), obesity, type 2 diabetes and others [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018].  However, the field's use of ML lacks clarity and consistency in which methods are used and how these methods are implemented [@thaiss_persistent_2016; @dadkhah_gut_2019]. More notably, we commonly see flawed ML practices such as (1) using ML pipelines where there is no seperate held-out test dataset to evaluate model performance or (2) reporting few or only the best outcomes of cross-validation. Even when there are seperate testing sets to evaluate model performance, there are large differences between cross-validation and testing performances that indicate overfitting as well as large confidence intervals for testing performances [@flemer_oral_2018; @baxter_microbiota-based_2016; @dai_multi-cohort_2018; @montassier_pretreatment_2016; @papa_non-invasive_2012; @mossotto_classification_2017; @ai_systematic_2017; @wong_quantitation_2017].

Moreover, there is a lack of discussion on why a particular ML model is utilized. Recently, there is a trend towards using more complex ML models such as random forest, extreme gradient boosting and neural networks without a discussion on if and how much model interpretibility is necessary for the study [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017; @ @geman_deep_2018]. Black box machine learning models are not inherently interpretable and require posthoc explanations to determine the feature importances in making a prediction. These explanations can be misleading and at times unreliable when making high-stake decisions about patient health [@rudin_please_2018]. The models we develop for healthcare, both to predict disease and to understand underlying reasons behind that prediction, should be transparent and accountable [@rudin_optimized_2018].

The lack of transparency on model selection and interpretation as well as flawed modeling methods negatively impact model  validity and reproducibiliy. We need to strive toward better machine learning practices by (1) implementing rigourous machine learning pipelines and (2) selecting ML models that reflect the goal of the study as it will inform our expectations of model accuracy, complexity, interpretibility and computational efficiency.To showcase a rigorous ML pipeline and to shed light on how ML model selection can affect modeling results, we performed an empirical analysis comparing several different ML models using the same dataset and the same ML pipeline. We used a previously published colorectal cancer (CRC) study [@baxter_dna_2016] which had fecal 16S rRNA gene sequences of `r paste(nrow(shared))` patients. The human-associated human microbiome is hypothesized to directly contribute to the development of CRC and fecal 16S rRNA gene sequences have been used to detect it [@baxter_microbiota-based_2016; @baxter_dna_2016; @zeller_potential_2014; @knights_human-associated_2011]. We built seven ML models using fecal 16S rRNA gene sequences to predict healthy patients versus patients with colorectal lesions that were identified by colonoscopy as screen relevant neoplasias (SRN). The study had `r paste(sum(sum(meta$dx=="normal")+sum(meta$Dx_Bin=="Adenoma")))` normal and `r paste(sum(sum(meta$dx=="cancer")+sum(meta$Dx_Bin=="adv Adenoma")))` SRN samples. We established modeling pipelines for L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost. We used the area under the receiver operating characteristic curve (AUROC) as a metric for determining the predictibve performance of each model. The AUROC ranges from 1.0, where the model perfectly distinguishes between cases and controls, to 0.50, where the model's predictions are no different from random chance. The median test AUROC varied from `r paste(round(test_all[performance_index[1],3], 3))`  to `r paste(round(test_all[performance_index[7],3], 3))`. Random forest had the highest median AUROC for detecting SRN. Despite its simplicity, the L2-regularized logistic regression was second best in predictive performance. In terms of computational efficiency, L1 SVM with linear kernel trained the fastest (`r paste(round(traintime_df[traintime_index[1],2], 3))` hours, std ± `r paste(round(traintime_df[traintime_index[1],3], 3))`), while XGBoost took the longest (`r paste(round(traintime_df[traintime_index[7],2], 3))` hours, std ± `r paste(round(traintime_df[traintime_index[7],3], 3))`). We also found that depending on how the data is split to create a held-out test set, the AUROC values of a ML model can vary up to `r paste(round(max((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 3))` which highlighted the importance of performing many randomized modeling runs. This study established standards for microbiome-associated ML models and underscored the importance of model selection.

@@

We used a cohort of 490 patients with 261 cases of SRN. For each patient, we had `r paste(ncol(data)-1)` features (fecal bacterial abundances) and a two-class label that defines their colorectal health (having colorectal lesions that are identified as SRN or normal). All the cases were independently labeled through colonoscopies. We established modeling pipelines for a binary prediction task with L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and extreme gradient boosted decision tree (XGBoost) to emphasize the differences in  model accuracy, complexity, interpretibility and computational efficiency due to  model selection.

For regularized logistic regression and SVM with linear kernel we used L2 regularization to keep all potentially important features. For comparison, we also trained an L1 regularized SVM model with linear kernel. L1-regularization on microbiome data lead to a sparser solution (i.e., force many coefficients to zero). Finally, to explore the potential for non-linear relationships among features and the outcome of interest, we trained tree based models, decision tree, random forest and XGboost, as well as an SVM with non-linear kernel.

We established a ML pipeline where we train and validate each of the seven models [Figure 1]. We randomly split the data into training/validation and test sets so that the training/validation set consisted of 80% of the full dataset while the test set was composed of the remaining data [Figure 1]. Since the cases are not uniformly represented in the data, the initial data-split was stratified to maintain the overall label distribution in both the training/validation and test sets. Training/validation set consisted of 393 patients (209 SRN), while the test set was composed of 97 patients (52 SRN). The training/validation data was used for training purposes and validation of hyperparameter selection (i.e. model selection), and the test set was used for evaluation purposes. Validation of hyperparameter selection was performed using repeated five-fold cross-validation on the training/validation set [Figure 1]. Similar to the initial data-split, five-fold cross-validation was also stratified to maintain the overall label distribution on the training and validation sets. We validated the cross-validation performances of each hyperparameter setting over 100 randomizations and selected the best performing hyperparameter setting in terms of AUROC metric and trained the full training/validation dataset [Figures S1 and S2]. We then used the held-out test set to evaluate the prediction performance of each ML model. The data-split, hyperparameter selection, training and testing steps were repeated 100 times to get a reliable and robust reading of model performance [Figure 1].

__Discriminative performance and generalizability of the seven models.__

We evaluated the predictive performances of seven binary classification models when applied to held-out test data using AUROC metric [Figure 2]. Random forest had significantly higher test AUROC values than the other models for detecting SRNs when AUROC values were compared to the other six by Wilcoxon rank sum test (p < 0.01). The median AUROC of the random forest model was `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Random_Forest"))$test_aucs), 3))`). L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel AUROC values were not significantly different from one another. They had median AUROC values of `r paste(round(median((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="XGBoost"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="XGBoost"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))`) and  `r paste(round(median((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))`) respectively. L1 SVM with linear kernel and decision tree had significantly lower AUROC values than the other ML models with median AUROC of `r paste(round(median((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3))`) and `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))`), respectively [Figure 2].

For each model, we compared the median cross-validation AUROC to the median testing AUROC. The difference between the two should be low to suggest the model is not overfitting despite the large number of features. The largest difference between the two was `r paste(round(abs(difference_model[difference_index[7],1]), 3))` in L1 SVM with linear kernel,  followed by SVM with radial basis function kernel and decision tree with a difference of `r paste(round(abs(difference_model[difference_index[6],1]), 3))` and `r paste(round(abs(difference_model[difference_index[5],1]), 3))`, respectively [Figure 2].

We reported the testing AUROC values over 100 randomizations of the initial data-split. The results showed that depending on the data-split, the testing AUROC values showed great variability [Figure 2]. The testing AUROC values within each model varied `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 2))` on average across the seven models. For instance, the lowest AUROC value of the random forest model was `r paste(round(min((all %>% filter(model=="Random_Forest"))$test_aucs), 2))` whereas the highest was `r paste(round(max((all %>% filter(model=="Random_Forest"))$test_aucs), 2))`.

@@

The ML models we built using L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost decrease in interpretibility as they increase in complexity. We interpreted L1 and L2 SVM with linear kernel and L2 logistic regression using the absolute feature weights of the trained models. We ranked the absolute weights of all the OTUs for each data-split [Figure 3]. We calculated the median ranks of these features over the 100 data-splits. In the three linear models, OTUs that had the largest median ranks and drove the detection of SRNs belonged to families *Lachnospiraceae*,  and *Ruminococcaceae* (OTU01239, OTU00659, OTU00742, OTU00012, OTU00015, OTU00768, OTU00822, OTU00609), genera *Gamella*  (OTU00426) and genera *Peptostreptococcus* (OTU00367) [Figure 3]. Some of the OTUs with the highest ranks were shared among the linear models.

We explained the feature importances in non-linear models; SVM with radial basis kernel, decision tree, random forest and XGBoost, using a method called permutation importance on the held-out test data.  Permutation importance analysis is where we randomly permute non-correlated features individually and groups of highly correlated features together. We then calculate how much the predictive performance of the model (i.e AUROC values) decrease when each OTU or group of OTUs is permuted randomly. We ranked the OTUs based on how much they decreased the median testing AUROC; the OTU with the largest decrease ranking highest. The top 5 OTUs with the largest negative impact on testing AUROC overlapped in tree-based models [Figure 4]. Specifically, permuting *Peptostreptococcus* (OTU00367) abundances randomly, dropped the predictive performances the most in all tree-based methods [Figure 4]. Decision tree, random forest and XGBoost models' predictive performance dropped from `r paste(round(dt_median, 2))` median AUROC to `r paste(round(median((dt_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, from `r paste(round(rf_median, 2))` to `r paste(round(median((rf_imp %>% filter(names=="Otu00367"))$new_auc), 2))` and from `r paste(round(xgboost_median, 2))` to `r paste(round(median((xgboost_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, respectively [Figure 4].

To highlight the differences between the two interpretation methods, we used permutation importance to interpret linear models as well [Figure S3]. L1-regularized SVM with linear kernel picked out some of the same OTUs (OTU00822, OTU01239, OTU00609) as important in the two interpretation methods; feature rankings based on weights [Figure 3] and permutation importance [Figure S3]. Similarly,  L2-regularized SVM and L2-regularized logistic regression picked out some of the same OTUs in both interpretation methods, OTU00659 and OTU00012, respectively. However, for all the linear models, the rankings of these features were different due to the collinearity in microbial communities.

@@

As the complexity of a ML model [Table S1] and the number of tuned hyperparameter settings increased [Figures S1-S2], its training times increased as well [Figure 5]. Linear models trained faster than non-linear models. L1 and L2 SVM with linear kernel and L2 logistic regression had training times of `r paste(round(traintime_df[traintime_index[1],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[1],3], 2))`), `r paste(round(traintime_df[traintime_index[2],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[2],3],2))`), and `r paste(round(traintime_df[traintime_index[3],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[3],3], 2))`), respectively. Whereas, a decision tree, SVM with radial basis function kernel, random forest and xgboost had training times of `r paste(round(traintime_df[traintime_index[4],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[4],3], 1))`), `r paste(round(traintime_df[traintime_index[5],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[5],3], 1))`), `r paste(round(traintime_df[traintime_index[6],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[6],3], 1))`) and `r paste(round(traintime_df[traintime_index[7],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[7],3], 1))`), respectively [Figure 5].

@@

  The colorectal health of the patient was defined as two encompassing classes; Normal or Screen Relevant Neoplasias (SRNs). Normal class includes patients with non-advanced adenomas or normal colons whereas SRN class includes patients with advanced adenomas or carcinomas. The bacterial abundances are the features used to predict colonic health of the patients. Bacterial abundances are discrete data in the form of Operational Taxonomic Unit (OTU) counts. OTU counts were set to the size of our smallest sample and were subsampled at the same distances. They were then transformd by scaling to a [0-1] range.

@@

**Permutation importance workflow.**
We created a Spearman's rank-order correlation matrix, corrected for multiple pairwise comparisons. We then defined correlated OTUs as having perfect correlation (correlation coef=1 and p<0.01).Other OTUs were permuted individually to get permutation importance but the correlated ones are grouped together and permuted at the same time. The reported OTUs that have impact on testing AUROC are reported in Figures xxx. If we want we can decrease the correlation coefficient to consider OTUs that are correlated with ecological consequences but this was out of the scope of this study but will be followed up in further analyses.

**Statistical analysis workflow.** Data summaries, statistical analysis, and data visualizations were performed using R (v.3.5.0) with the tidyverse package (v.1.2.1). We compared the AUROC values of the seven ML models by Wilcoxon rank sum tests to determine the best discriminative performance.

>>>>

---
title: '**Best practices for applying machine learning to bacterial 16S rRNA gene sequencing data**'
csl: mbio.csl
fontsize: 11pt
output:
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
  word_document: default
geometry: margin=1.0in
bibliography: references.bib
---

@@

Begüm D. Topçuoğlu${^1}$, Nicholas A. Lesniak${^1}$, Jenna Wiens${^2}$, Mack Ruffin${^3}$, Patrick D. Schloss^1$\dagger$^

@@

Machine learning (ML) modeling of the human microbiome has the potential to identify the microbial biomarkers and aid in diagnosis of many chronic diseases such as inflammatory bowel disease, diabetes and colorectal cancer. Progress has been made towards developing ML models that predict health outcomes from bacterial abundances, but rigourous ML models are scarce due to the flawed methods that call the validity of developed ML models into question. Furthermore, the use of black box ML models has hindered the validation of microbial biomarkers. To overcome these challenges, we benchmarked seven different ML models that use fecal 16S rRNA sequences to predict colorectal cancer (CRC) lesions (n=490 patients, 261 controls and 229 cases). To show the effect of model selection, we assessed the predictive performance, interpretability, and computational efficiency of the following models: L2-regularized logistic regression, L1 and L2-regularized support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest, and extreme gradient boosting (XGBoost). The random forest model was best at detecting CRC lesions with an AUROC of 0.695 but it was slow to train (83.2 h) and hard to interpret. Despite its simplicity, L2-regularized logistic regression followed random forest in predictive performance with an AUROC of 0.680, and it trained much faster (12 min). In this study, we established standards for the development of modeling pipelines for microbiome-associated ML models. Additionally, we showed that ML models should be chosen based on expectations of predictive performance, interpretability and available computational resources.

@@

Prediction of health outcomes using ML is rapidly being adopted by human microbiome studies. However, the developed ML models so far are overoptimistic in terms of validity and predictive performance. Without rigorous ML pipelines, we cannot trust ML models. Before we can speed up progress, we need to slow down, define and implement good ML practices.
\newpage

@@

Advances in sequencing technology and decreases in the costs of generating 16S rRNA gene sequences have allowed rapid exploration of the human microbiome and its health implications. Currently, the human microbiome field is growing at an unprecedented rate and there is an increasing demand for methods that identify associations between microbiome members and human health. However, this is difficult as human associated microbial communities are remarkably complex, high-dimensional and uneven within and between individuals with the same disease. It is unlikely that a single species can explain a disease. Instead, subsets of those communities, in relation to one another and to their host, account for differences in health outcomes.

Machine learning (ML) methods are effective at recognizing and highlighting patterns in complex microbial datasets. They learn from existing data to predict the outcomes of new data and allow us to infer on the reasons underlying that prediction. Therefore, researchers have started to explore the utility of ML models that use microbiota associated biomarkers to predict human health and to understand the ecological basis of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases (IBD), obesity, and type 2 diabetes [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018]. The high-stake task of predicting the correct diagnosis with high confidence relies on a ML model that is built with rigorous methods. However, the field's use of ML lacks clarity and consistency in which methods are used and how these methods are implemented [@thaiss_persistent_2016; @dadkhah_gut_2019]. More notably, we commonly see flawed ML practices such as using ML pipelines where there is no seperate held-out test dataset to evaluate predictive performance or reporting few or only the best outcomes of cross-validation. Even when there are seperate testing sets to evaluate predictive performance, there are large differences between cross-validation and testing performances and large confidence intervals in testing performances [@flemer_oral_2018; @baxter_microbiota-based_2016; @dai_multi-cohort_2018; @montassier_pretreatment_2016; @papa_non-invasive_2012; @mossotto_classification_2017; @ai_systematic_2017; @wong_quantitation_2017]. These practices prevent the development of generalizable models, where the model makes accurate predictions with newly acquired data as well as it does with the training data.

Moreover, there is a lack of discussion on why a particular ML model is utilized and a lack of emphasis on the strengths and weaknesses of it. Recently, there is a trend towards using more complex ML models such as random forest, extreme gradient boosting (XGBoost) and neural networks without a discussion on if and how much model interpretibility is necessary for the study [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017; @ @geman_deep_2018]. These models are also called black box ML models because they are not inherently interpretable and require posthoc explanations to determine the feature importances in making a prediction. These explanations can be misleading and at times unreliable when making high-stake decisions about patient health [@rudin_please_2018]. The models we develop for healthcare, both to predict disease and to understand underlying reasons behind that prediction, should be transparent and accountable [@rudin_optimized_2018].

The lack of transparency on model selection and interpretation as well as flawed modeling methods negatively impact model  validity and reproducibility. We need to strive toward better machine learning practices by (1) implementing rigorous machine learning pipelines and (2) selecting ML models that reflect the goal of the study as it will inform our expectations of predictive performance, complexity, interpretability and computational efficiency. To showcase a rigorous ML pipeline and to shed light on how ML model selection can affect modeling results, we performed an empirical analysis comparing several different ML models using the same dataset and the same ML pipeline. We used a previously published colorectal cancer (CRC) study [@baxter_dna_2016] which had fecal 16S rRNA gene sequences of CRC patients. The human microbiome is hypothesized to directly contribute to the development of CRC and fecal 16S rRNA gene sequences have been used to detect it [@baxter_microbiota-based_2016; @baxter_dna_2016; @zeller_potential_2014; @knights_human-associated_2011]. We built seven ML models using fecal 16S rRNA gene sequences to predict healthy patients versus patients with colorectal lesions that were identified by colonoscopy as screen relevant neoplasias (SRN). We established modeling pipelines for three linear models with different forms of regularization; L2-regularized logistic regression, L1 and L2-regularized support vector machines (SVM) with linear kernel. We also developed four non-linear models; SVM with radial basis function kernel, a decision tree, random forest and XGBoost. We compared the predictive performance, interpretability and computational efficiency of the seven models to highlight the importance of model selection. We established standards for ML pipeline construction, predictive performance evaluation and model interpretation for microbiome-associated ML models that have a binary prediction task.

@@

We first determined the dataset and the ML models to use in our study. We used a previously published study on a CRC cohort of 490 patients with 261 cases of SRN. For each patient, we had `r paste(ncol(data)-1)` features (fecal bacterial abundances) and a two-class label that defines their colorectal health (normal or SRN colorectal lesions as defined by colonoscopies). We established modeling pipelines for a binary prediction task with L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and extreme gradient boosted decision tree (XGBoost).

We used ML models with different classification algorithms and regularization parameters. For regularized logistic regression and SVM with linear kernel, we used L2 regularization to keep all potentially important features. For comparison, we also trained an L1 regularized SVM model with linear kernel. L1-regularization on microbiome data lead to a sparser solution (i.e., force many coefficients to zero). Finally, to explore the potential for non-linear relationships among features and the outcome of interest, we trained tree based models, decision tree, random forest and XGBoost, as well as an SVM with non-linear kernel.

We established a ML pipeline where we train and validate each of the seven models [Figure 1]. We randomly split the data into training/validation and test sets so that the training/validation set consisted of 80% of the full dataset while the test set was composed of the remaining data [Figure 1]. Since the cases are not uniformly represented in the data, the initial data-split was stratified to maintain the overall label distribution in both the training/validation and test sets. Training/validation set consisted of 393 patients (209 SRN), while the test set was composed of 97 patients (52 SRN). The training/validation data was used for training purposes and validation of hyperparameter selection, and the test set was used for evaluation purposes.

Hyperparameters are the rules that are learned from the training data in a classification algorithm. When they are tuned over a full grid search and selected by validation, they make the ML model better are predicting. We selected hyperparameter settings using repeated five-fold cross-validation on the training/validation set [Figure 1]. We chose the best hyperparameter setting for each model based on its predictive performance on the validation set using the area under the receiver operating characteristic curve (AUROC) metric [Figure S1 and S2]. The AUROC ranges from 1.0, where the model perfectly distinguishes between cases and controls, to 0.50, where the model's predictions are no different from random chance. Similar to the initial data-split, five-fold cross-validation for hyperparameter selection was also stratified to maintain the overall label distribution on the training and validation sets. The cross-validation of each hyperparameter setting was repeated over 100 randomizations to get a robust reading of predictive performance.

We then trained the full training/validation dataset with the selected hyperparameters. We used the held-out test set to evaluate the predictive performance of each ML model. The data-split, hyperparameter selection, training and testing steps were repeated 100 times to get a reliable and robust reading of model performance [Figure 1].

__Predictive performance and generalizability of the seven models.__

We evaluated the predictive performances of seven binary classification models when applied to held-out test data using the AUROC metric [Figure 2]. Random forest had significantly higher test AUROC values than the other models for detecting SRNs when AUROC values were compared to the other six by Wilcoxon rank sum test (p < 0.01). The median AUROC of the random forest model was `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Random_Forest"))$test_aucs), 3))`). L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel AUROC values were not significantly different from one another. They had median AUROC values of `r paste(round(median((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="XGBoost"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="XGBoost"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))`) and  `r paste(round(median((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))`) respectively. L1-regularized SVM with linear kernel and decision tree had significantly lower AUROC values than the other ML models with median AUROC of `r paste(round(median((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3))`) and `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))`), respectively [Figure 2]. Random forest had the highest median AUROC for detecting SRN. Despite its simplicity, the L2-regularized logistic regression was second best in predictive performance.

To evaluate the generalizability of each model, we compared the median cross-validation AUROC to the median testing AUROC. The difference between the two should be low to suggest the model is not overfitting despite the large number of features. The largest difference between the two was `r paste(round(abs(difference_model[difference_index[7],1]), 3))` in L1-regularized SVM with linear kernel, followed by SVM with radial basis function kernel and decision tree with a difference of `r paste(round(abs(difference_model[difference_index[6],1]), 3))` and `r paste(round(abs(difference_model[difference_index[5],1]), 3))`, respectively [Figure 2]. We also reported the testing AUROC values over 100 randomizations of the initial data-split. The testing AUROC values within each model varied `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 2))` on average across the seven models. For instance, the lowest AUROC value of the random forest model was `r paste(round(min((all %>% filter(model=="Random_Forest"))$test_aucs), 2))` whereas the highest was `r paste(round(max((all %>% filter(model=="Random_Forest"))$test_aucs), 2))`. These results showed that depending on the data-split, the testing AUROC values showed great variability [Figure 2].

@@

Interpretability is the degree to which humans can understand the reasons behind a model prediction [@miller_explanation_2017]. Because we often use ML models not just to predict a health outcome but also to learn the ecology behind a disease, model interpretation becomes crucial for microbiome studies. The ML models we built using L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost decrease in interpretability as they increase in complexity.  In this study we highlighted two methods to interpret models with varying complexity.

We interpreted linear models (L1 and L2-regularized SVM with linear kernel and L2-regularized logistic regression) using the absolute feature weights of the trained models. We ranked the absolute weights of all the OTUs for each data-split [Figure 3]. We calculated the median ranks of these features over the 100 data-splits. In the three linear models, OTUs that had the largest median ranks and drove the detection of SRNs belonged to families *Lachnospiraceae*,  and *Ruminococcaceae* (OTU01239, OTU00659, OTU00742, OTU00012, OTU00015, OTU00768, OTU00822, OTU00609), genera *Gamella*  (OTU00426) and genera *Peptostreptococcus* (OTU00367) [Figure 3]. Some of the OTUs with the highest ranks were shared among the linear models.

We explained the feature importances in non-linear models; SVM with radial basis kernel, decision tree, random forest and XGBoost, using a method called permutation importance on the held-out test data.  Permutation importance analysis is a posthoc explanation of the model where we randomly permute non-correlated features individually and groups of highly correlated features together. We then calculate how much the predictive performance of the model (i.e AUROC values) decrease when each OTU or group of OTUs is permuted randomly. We ranked the OTUs based on how much they decreased the median testing AUROC; the OTU with the largest decrease ranking highest. The top 5 OTUs with the largest negative impact on testing AUROC overlapped in tree-based models [Figure 4]. Specifically, permuting *Peptostreptococcus* (OTU00367) abundances randomly, dropped the predictive performances the most in all tree-based methods [Figure 4]. Decision tree, random forest and XGBoost models' predictive performance dropped from `r paste(round(dt_median, 2))` median AUROC to `r paste(round(median((dt_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, from `r paste(round(rf_median, 2))` to `r paste(round(median((rf_imp %>% filter(names=="Otu00367"))$new_auc), 2))` and from `r paste(round(xgboost_median, 2))` to `r paste(round(median((xgboost_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, respectively [Figure 4].

To highlight the differences between the two interpretation methods, we used permutation importance to interpret linear models as well [Figure S3]. L1-regularized SVM with linear kernel picked out some of the same OTUs (OTU00822, OTU01239, OTU00609) as important in feature rankings based on weights [Figure 3] and permutation importance [Figure S3]. Similarly, L2-regularized SVM and L2-regularized logistic regression picked out some of the same OTUs in both interpretation methods, OTU00659 and OTU00012, respectively. However, for all the linear models, the rankings of these features were different due to the collinearity in microbial communities.

@@

We compared the training times of the seven ML models. As the complexity of a ML model [Table S1] and the number of tuned hyperparameter settings increased [Figures S1-S2], its training times increased as well [Figure 5]. Linear models trained faster than non-linear models. L1 and L2 SVM with linear kernel and L2-regularized logistic regression had the shortest training times with `r paste(round(traintime_df[traintime_index[1],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[1],3], 2))`), `r paste(round(traintime_df[traintime_index[2],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[2],3],2))`), and `r paste(round(traintime_df[traintime_index[3],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[3],3], 2))`), respectively. Whereas, a decision tree, SVM with radial basis function kernel, random forest and XGBoost had training times of `r paste(round(traintime_df[traintime_index[4],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[4],3], 1))`), `r paste(round(traintime_df[traintime_index[5],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[5],3], 1))`), `r paste(round(traintime_df[traintime_index[6],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[6],3], 1))`) and `r paste(round(traintime_df[traintime_index[7],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[7],3], 1))`), respectively [Figure 5].

@@

  The colorectal health of the patient was defined as two encompassing classes; Normal or Screen Relevant Neoplasias (SRNs). Normal class includes patients with non-advanced adenomas or normal colons whereas SRN class includes patients with advanced adenomas or carcinomas. The study had 261 normal and 229 SRN samples. The bacterial abundances are the features used to predict colonic health of the patients. Bacterial abundances are discrete data in the form of Operational Taxonomic Unit (OTU) counts. OTU counts were set to the size of our smallest sample and were subsampled at the same distances. They were then transformed by scaling to a [0-1] range.

@@

**Permutation importance workflow.**
We created a Spearman's rank-order correlation matrix, corrected for multiple pairwise comparisons. We then defined correlated OTUs as having perfect correlation (correlation coef=1 and p<0.01). Non-correlated OTUs were permuted individually whereas correlated ones were grouped together and permuted at the same time.

**Statistical analysis workflow.** Data summaries, statistical analysis, and data visualizations were performed using R (v.3.5.0) with the tidyverse package (v.1.2.1). We compared the AUROC values of the seven ML models by Wilcoxon rank sum tests to determine the best predictive performance.
